But I can't preregister my research: Improving the reproducibility and transparency of ecology and conservation with adaptive preregistration for model-based research

### 2024-04-30 _analysis/ms.qmd
First Draft Manuscript - add content from draft obsidian doc - mv elsevier quarto extension - commit PDF in case there are rendering issues down the line :|

  **1.** Preregistration is an open-science practice adopted in scientific disciplines that have begun to confront the 'reproducibility crisis', which aims to improve research transparency and mitigate questionable research practices within a study. There has been little uptake of preregistration in ecology and conservation, and many modellers and researchers engaged in non-hypothesis testing research both within and outside of ecology have eschewed preregistration on the grounds that existing templates are irrelevant because they are focussed on null-hypothesis significance testing.
  **2.** We advocate that preregistration can and should be used in model-based research in ecology and conservation but that preregistration should reflect the norms and practice of the research context in which it is applied in order to adequately restrict researcher degrees of freedom that are unique to a domain-specific methodology. We took a user-centred design approach to the task of translating preregistration into model-based research contexts for ecology and conservation.
  **3.** We identified a generalised 'modelling workflow' that captures critical analysis decisions in the process of the model development cycle. The workflow informed the content and structure of a preregistration template to be used in model-based research. The iterative nature of model development was identified as a key barrier to implementing preregistration for modelling research because it conflicts with the current format and linear process of preregistration. To better align the internal logic of preregistration with the iterative and non-linear process of ecological modelling, we propose a methodology for implementing 'adaptive preregistration' leveraging git and GitHub. We tested and evaluated the preregistration template and methodology using a case-study of environmental flows modelling for decision-support in Victoria, Australia.
  **4.** This research provides a template and methodology for implementing adaptive preregistration of ecological models in either pure or applied settings, and will facilitate the wider uptake of preregistration in ecology. Although we focus on ecology and conservation in this paper, the template is extensible to alternative fields and methods, and the proposed methodology for implementing adaptive preregistration can be applied in model-based research within science more broadly. Modellers need no longer cry "but, I can't preregister my research."

@@

@@

# Materials and Methods

@@

# Discussion & Conclusions

# Data & Software Availability

# Author Contributions

# Supplementary Material

>>>>

  1. Preregistration is an open-science practice adopted in scientific disciplines that have begun to confront the ‘reproducibility crisis’, which aims to improve research transparency and mitigate questionable research practices within a study. There has been little uptake of preregistration in ecology and conservation, and many modellers and researchers engaged in non-hypothesis testing research both within and outside of ecology have eschewed pre- registration on the grounds that existing templates are irrelevant because they are focused on null-hypothesis significance testing.
  2. We advocate that preregistration can and should be used in model-based research in ecology and conservation, but that preregistration should reflect the norms and practice of the research context in which it is applied, in order to adequately restrict researcher degrees of freedom that are unique to a domain-specific methodology. We took a user-centred design approach to the task of translating preregistration into model- based research contexts for ecology and conservation.
  3. We identified a generalised ‘modelling workflow’ that captures critical analysis decisions in the process of the model development cycle. The workflow informed the content and structure of a preregistration template to be used in model-based research. The iterative nature of model development was identified as a key barrier to implementing preregistration for modelling research because it conflicts with the current format and linear process of preregistration. To better align the internal logic of preregistration with the iterative and non-linear process of ecological modelling, we propose a methodology for implementing ‘adaptive preregistration’ leveraging git and GitHub. We tested and evaluated the preregistration template and methodology using a case- study of environmental flows modelling for decision-support in Victoria, Australia.
  4. This research provides a template and methodology for implementing adaptive preregistration of ecological models in either pure or applied settings and will facilitate the wider uptake of preregistration in ecology. Although we focus on ecology and conservation in this paper, the template is extensible to alternative fields and methods, and the proposed methodology for implementing adaptive preregistration can be applied in model- based re- search within science more broadly. Modellers need no longer cry “but, I can’t preregister my research.”

@@

bibliography: references_obsidian.bib

@@

Science's replication crisis, and the subsequent recognition of problems with reliability of scientific research findings, has triggered a 'credibility revolution' spawning a suite of proposed methodological and policy reforms, including *preregistration* [@parker2019]. Preregistration is an open-science practice that aims to distinguish between planned and ad hoc analyses  [@Parker2019] and aims to combat the risk of questionable research practices by requiring researchers to register their methods and analysis plan in a secure and publicly accessible platform prior to collecting and/or analysing data, which cannot be altered after submission ( #todo #citation). The purpose of preregistration is to prevent biased research findings by guarding against questionable research practices (such as selective reporting, p-hacking, HARKing, among others #todo #citation ) by distinguishing what is a genuine *a priori* planned analysis, and what is not, thereby preventing the opportunistic misuse of researcher degrees of freedom, or those many analytic decisions researchers make during the planning, data collection, data analysis and reporting of results in scientific studies (See Box 1 for background). QRPs have been labelled as one of the primary cause of the replication crisis (Shrout and Rodgers, 2018 in [@Fife2022]) and serve to XYZ bias the overall literature, etc ( #todo finish this point).

Despite enthusiastic uptake in some scientific disciplines, such as psychology and medicine ( #todo #citation ), the use of registered reports and preregistration in ecology and conservation more broadly is new and limited [@Desbureaux2021], particularly for model-based research ( #todo Joel Pick preprint?)**. Ecologists have brushed preregistration off on the basis that existing templates are irrelevant since they are focussed on null-hypothesis significance testing**, whereas in applied ecology and conservation, we don't usually explicitly test hypotheses [@Betts2021], but instead use explanatory models, or increasingly predictive models that generate anticipatory predictions, to inform management decisions [@Wood2020]. Model-fitting and evaluation, rather than falsification of hypotheses about natural phenomena, are increasingly prevalent approaches to scientific inquiry in ecology [@Connolly2017]

Preregistration in model-based research has been eschewed on several grounds by modellers and researchers engaged in model-based research both within and outside of ecology. Much of the literature advocating for the use of preregistration assumes a hypothetico-deductive model of scientific inquiry, and the templates used to write preregistrations have been accused of being focussed on null-hypothesis significance testing [e.g. @McDermott2021; @Prosperi2019]. Approaches and advocacy for preregistration of MBR across all scientific disciplines are limited, and focussed have focused on the phases of the modelling process and modelling approaches that are more confirmatory ( #todo see @ioannidis2022). For example, @Cruwell2019b present a template for preregistration of MBR where the model construction and development phase is explicitly and intentionally excluded from the preregistration process, while much of the discussion around preregistration for models has centred on machine learning [@Hildebrandt2018] or predictive modelling [@hofman2023] ( #todo see \[preregistration for modelling\] for more refs).

A second argument against preregistration for model-based research is based on the mischaracterisation of the model development process as a intrinsically exploratory process [@McDermott2021] , and since only the preregistration of confirmatory research, such as hypothesis testing, is considered necessary [@Prosperi2019], there is no need to pre-register model-based research ( #todo #citation). The description of modelling as a wholly exploratory process is a mischaracterisation based on a misunderstanding of sensemaking and reasoning within the modelling process stemming from a lack of clear formal definitions distinguishing confirmatory research from exploratory research [@Flournoy2020; @Devezer2020; @alspaugh2019]. Although modelling includes some tasks and activities that are exploratory in nature, it can be distinguished from wholly *exploratory analysis* in that tends be goal-driven, directed analysis that is less open-ended [@alspaugh2019]. For model-based research, the exploratory/confirmatory divide is an idealised and simplifying dichotomy that, in practice, is more complicated and nuanced and can be considered as a spectrum on which exploratory analysis sits at the analysis is open-ended, high-level and opportunistic end, to goal-driven, low-level and precise analyses at the other [@alspaugh2019]. In practice the divide is often blurred, and researchers will move back and forth between different types of reasoning and inference throughout the modelling process, switching from activity at one end of the spectrum to the other and back again [@alspaugh2019], with deductions leading to abductions as the analysis process proceeds [@alspaugh2019; @Prosperi2019], or even occuring together in a hybridized manner [@Prosperi2019].

When critics of preregistration for MBR describe modelling as intrinsically 'exploratory', what is seemingly being referred to is the iterative, adaptive and non-linear nature of model-development. In practice, modelling proceeds in an adaptive manner by which models are iteratively developed and refined [@Augusiak:2014gz; @Babel2019; @Bennett2013; @Bodner2020]. New analyses are often chosen based on the basis of data exploration and previous analyses of the same data [@Dwork2015], For example, modellers may perform exploratory analyses to validate distributional assumptions are met, such as no over-dispersion or zero-inflation in count-data to ensure valid inference when using particular models [@Campbell2021] ( #todo also see @alspaugh2019 for list of examples). Such practices violate a key tenet of preregistration, that is 'data-independent' analytic decision-making ( #todo #citation). Consequently, the iterative nature of model development, wherein models are continually refined through analysing and reanalysing a dataset and the model, is considered to be in conflict with the practice of preregistration; by requiring researchers to commit to a single static plan that prevents changes or deviations from the original plan it is seen to be forcing adaptive analysis to be *non-adaptive* through 'defining the entire analysis protocol ahead of time' [@Dwork2015; @Benning2019; @MacEachern2019; @McDermott2021]. In short, we can't write off preregistration for modelling on the basis that it is exploratory, because in practice it employs a range of modes of reasoning from induction, deduction, to abduction, with analysts switching between those types of reasoning or even hybridising them.

What *is* important to consider when considering the application of preregistration within MBR is whether the reliability of MBR may be potentially threatened by the potential for Questionable Research Practices (Box 1). Evidence from ecology more broadly has revealed that the rate of QRPs is similar to those in other disciplines [@Fraser:2018cl] and while there has been little study of QRPs in MBR (except @Gould:2020ebg #todo preprint doc & cross-ref here), preliminary evidence reveals that modellers do engage in QRPs such as selective reporting, for example, they may may omit explored analysis paths from their reported analyses if they did not corroborate their conclusions [@Liu2020]. Furthermore, QRPs are likely to arise in conditions where there is incomplete and opaque reporting practices [@Fidler:2018ho], and poor model-reporting practices and an overall lack of transparency about the modelling process are endemic in MBR in ecology. Concerns about the state of model reporting in ecological modelling have been raised repeatedly, arguing that current practices are neither transparent or comprehensive [@Schmolke:2010fd; @deVos:2011uk]. Despite a number of attempts to promote thorough and transparent reporting, (e.g. @Jakeman:2006ii; @Alexandrov:2011iv; @Bodner2020) as a critical component of *Good Modelling Practice*, a culture of adequate research transparency in modelling is yet to be established and remains limited [@deVos:2011uk; @Grimm:2014es]. The lack of transparency in describing ecological models and reporting the modelling process and results provides significant opportunity for undisclosed researcher degrees of freedom, and hence, the potential for Questionable Research Practices.

Although the purpose of preregistration is predominantly conceptualised in terms of distinguishing exploratory from confirmatory research (e.g. @Nosek:2018eg, @munafo2017 and #todo insert other refs), we we adopt a more general stance towards the purpose of preregistration, i.e. that preregistration may mitigate against questionable research practices by reducing the potential for cognitive biases to cause opportunistic misuse of researcher degrees of freedom. In this paper, we advocate that preregistration can and should be used in model-based research in ecology and conservation, but its particular form should reflect the norms and practices of the research context in which it is applied, in order to adequately restrict researcher degrees of freedom that may uniquely emerge within this domain-specific methodology.

The goal of this study is to rethink preregistration to better suit the domain-specific challenges and methodological particularities of model-based research in ecology and conservation. Preregistration cannot simply be ported over to other disciplines and adopted as-is. Different research methodologies, and their domain-specific application across different domains and fields of study follow different analytical choices and are therefore susceptible to different QRPs. Consequently, domain-specific templates are necessary for guiding preregistration in such a way that restricts researcher degrees of freedom as they uniquely emerge within a domain or field-specific application of a particular research methodology. If we wish to adopt preregistration for ecological modelling, both the content required to be preregistered, as well as the procedure for creating a preregistration must be adapted to suit model-based research as it is undertaken in ecology.

We first aimed to build a preregistration template for ecological modelling. We identified a generalised ‘modelling workflow’ that captured critical analysis decisions in the process of the model development cycle. The workflow informed the content and structure of a preregistration template to be used in model-based research. The iterative nature of model development was identified as a key barrier to implementing preregistration for modelling research because it conflicts with the current format and linear process of preregistration. To better align the internal logic of preregistration with the iterative and non-linear process of ecological modelling, we devised a methodology for implementing an expanded view of preregistration that we term *Adaptive Preregistration* [sensu @Srivastava2018]. Adaptive preregistration allows for the iterative cycle of model development to be reflected in the preregistration and better accounts for the inherently adaptive nature of model development than current approaches to preregistration in which researchers are committed to a static, unalterable and deterministic view of preregistration. Finally, to assess the feasibility and practicality of preregistration for MBR, we tested and evaluated the preregistration template and methodology using a case-study of environmental flows modelling for decision-support in Victoria, Australia.

------------------------------------------------------------------------

# Box 1: How Researcher Degrees of Freedom become Questionable Research Practices in MBR

Modellers are faced with numerous 'researcher degrees of freedom' [@Wicherts2016] or subjective decisions during the course of modelling that may substantially impact the chosen analysis strategy, and therefore the research findings and conclusions drawn from that analysis [@Hoffmann2021; @Simonsohn2020; @Desbureaux2021; @gouldSameDataDifferent2023]. The set of all plausible or reasonable potential decisions that a researcher could take from the beginning to end of analysis has been described as the 'garden of forking paths' [@Gelman2013]. 'Many analyst'-style studies, whereby multiple separate teams of researchers independently analyse the same dataset and same research question, demonstrate that different pathways in this 'garden of forking paths' can lead to potentially conflicting conclusions [@Young2021; @Liu2020; @gouldSameDataDifferent2023].

Researcher degrees of freedom arise because there is inherent analytic uncertainty in model-based research whether due to incomplete theoretical understanding of underlying ecological processes, or because modelling is an act of simplification and abstraction; there is always some degree of choice about which ecological processes and phenomena should be included in the model, how they should be represented [@Babel2019; @Getz:2017ic]. Further analytic uncertainty may arise due to uncertainty in data processing or choice about which analysis method is best [@Hoffmann2021].

-   *Analytic Uncertainty* [@Walker2003] talks about uncertainty specifically in terms of *model based* decision-making. #todo

When researchers are faced with ambiguous decisions where there is no single decision or strategy acceptable according to scientific standards and consensus, we are likely to select the decision that results in the most 'favourable' finding with "convincing self-justification" [@Simmons:2011iw; @Hoffmann2021]. Researchers routinely execute and explore alternative analysis pathways in service of identifying the most favourable (i.e. publishable) result, selectively reporting only one or several preferred results [@Liu2020; @Simmons:2011iw]. Thus researcher degrees of freedom precipitate 'questionable' research practices, or QRPs, when they remain both undisclosed and exploited opportunistically [@Bakker2018]. Such 'questionable' research practices have been demonstrated to be surprisingly common upon admission among ecologists, with comparable rates to other scientific disciplines [@Fraser:2018cl]. Questionable research practices are categorically distinct from fraudulent practices, arising not from consciously nefarious 'fishing expedition\[s\]' [@Gelman2013], but rather because researchers are motivated actors working within a 'publish or perish' culture who are susceptible to cognitive biases and are willing to change their modelling assumptions when the results conflict with desired outcomes, for example, sincerely and self-convincingly believing that the more favourable model is superior [@Young2021].

QRPs may increase the probability of false positive results, may artificially inflate effect size estimates, and ultimately reduce the replicability of published findings, while encouraging overconfidence in the precision of results and exacerbating the risk of accepting and propagating false facts [@Wicherts2016; @Hoffmann2021; @Roettger2019; @Nissen2016]. Good intentions, and simply being 'aware' of the potential for researcher degrees of freedom to lead to questionable research practices cannot sufficiently mitigate their risk [@Nosek:2018eg].

------------------------------------------------------------------------

# Methods

We approached the task of translating preregistration into model-based research contexts for ecology and conservation with a user-centred design (UCD) framework. UCD is a philosophy for designing processes, services or products that account for user requirements at all stages of the design process so that the resource generated by this approach is effective and usable for specific people in specific contexts [@Pavelin2012]. Because it is difficult to predict how users will actually engage with a resource [@Pavelin2012] and a designer's intention for preregistration may conflict with the user's purpose and undermine the overarching goal of preregistration [@Pu2019], we heed the advice of Pu et al. [-@Pu2019], who recommend treating preregistration as a UCD problem so that preregistration can effectively meet its intended objectives, despite UCD being typically used within software engineering domains, rather than academic domains.

We borrow from the methodology proposed by Pavelin et al. [-@Pavelin2012] to design, test and refine a preregistration template suitable for ecological modelling, and a methodology for preregistering model-based research (@fig-UCD). The methods description is structured following the key steps illustrated in @fig-UCD.

![Insert caption. Number each step so i can refer to it in the headings and text below. Make it clear in the figure that we're not doing the build interface and conduct usability testing component of implement, nor the evaluate and iterate. we throw that out to the community of practice to do that part.](images/Pasted image 20240314123346.png){#fig-UCD}

## Define Vision & Objectives

> Define the goals of the research under development and identify the audience

The first step in a UCD approach is to articulate the goals, objectives and target audience of the resource, consequently we define the intended purpose of the preregistration template and methodology, the community of users, as well as their research contexts and use-cases. Finally, we explicate the aspirations and desirable attributes of the template and methodology. Much like in 'Good Modelling Practice', articulating the overarching purpose, intended audience, as well as the objectives and aspirations will guide the development and evaluation of preregistration for model-based research in ecology and conservation, with important implications for the design of the template -- from its structure, content, user-interface and broader workflow in which it is designed to be used.

### Articulating Intended purpose of preregistration template & methodology

We define the purposes of the template as:

1\. *To delimit researcher degrees of freedom* \> when conducting model-based research in applied ecology and conservation so as to mitigate the risk of questionable research practices in order to improve the reliability and credibility of this research.

2\. *Increase research transparency* \> of model-based research in ecology and conservation. *Research transparency* refers to both production transparency — which includes research artefacts like open-access data and materials, or data collection procedures — and analytic transparency — which describes a complete account of how the analytic conclusions are drawn from the data [@Lupia2014].

### Target user community and their research-contexts

We aim for the template and preregistration process developed in this study to be used for model-based research, which we define as any research, empirical or applied, that uses quantitative modelling to answer its research question [@Brudvig2017], where *quantitative modelling* may be defined as any modelling that "describes and/or forecast(s) the behaviour of a system using mathematical and statistical concepts, and whose parameters and their relationships are expressed as quantities" [@Garcia-Diaz2019].

We define model-based research loosely in order to capture the breadth of different modelling goals (e.g. exploration, inference and prediction; @Tredennick2021) and type of research contributions (e.g. development of a new method, etc #todo see @Moallemi2019 and across both pure and applied ecological modelling.

### Specifying Preregistration use-cases

Pu et al. [-@Pu2019] describe two key use-cases of preregistration, depending on the role assumed in interacting with the preregistration. Users have different requirements of preregistration depending on the role that they assume, that is, study authors and modellers have a different use-case than reviewers, editors and readers, however this latter use-case is often not explicitly designed and accounted for in current platforms/infrastructure/processes [@Pu2019]. Reviewers and editors are usually tasked with the process of 'checking' the preregistration, comparing the final manuscript to its preregistration and verifying that the preregistered analyses are conducted and reported as specified [@Pu2019]. However existing formats are designed primarily for the use-case of the author - they are text-based static documents that are ill-suited to checking the reported study against the preregistration.

Although preregistration authorship is the primary target use-case of the preregistration template and methodology developed in this study, we aim to consider and incorporate features into the preregistration process, such as template formatting, that will facilitate preregistration checking to also address non-authorship use-cases.

### Articulating desirable attributes of the preregistration template and adaptive preregistration methodology

The preregistration template should capture critical decision-points that may influence the overall model and analysis as well as any recommendations or decisions drawn from analysis of the model, thus delimiting researcher degrees of freedom at these junctures in the modelling decision process. The structure and content of the template is necessarily opinionated, encouraging Good Modelling Practices, while reflecting community norms and practices in model-development while still meeting the objective of mitigating the risk of Questionable Research Practices by delimiting researcher degrees of freedom. The template needs to be generalisable enough to apply to a range of common modelling purposes, methods and research contributions, and flexible enough that if a particular decision is not applicable for the planned study, that those items may be removed or altered, and other items may be added onto the template.

Secondary to the aim of mitigating QRPs, is that the the template should also enhance the research transparency of the preregistered study, ensuring that analysis choices are documented with sufficient transparency to facilitate the reader of the preregistered completed study in making informed assessments about the credibility and reliability of the model and any associated analysis and recommended interventions. Borrowing from @Bodner2020's framework to increase the accuracy, reliability and transparency of predictive models, in order to increase transparency of model-reporting, the template should explicitly outline critical model choices and their assumptions, steps in the modelling process, and explicit expectations about the outputs and analysis of the model.

**Adaptive Preregistration Methodology**

The Adaptive Preregistration Methodology aims to capture the entire modelling process, including model development and construction, rather than beginning at the point of analysing a pre-existing and parameterised model for deductive inference / prediction. Critically, the methodology must not preclude the iterative and adaptive nature of model construction, while simultaneously restricting the types of data-dependent analytic decision-making that may compromise the reliability and validity of the model and its contingent analysis and conclusions.

## User Research & Analysis

> Characterise users and their needs

On the 4th and 5th of March 2020, we facilitated a series of structured activities at the Arthur Rylah Institute for Environmental Research which aimed to inform the development of two preregistration templates, one for field-work and sampling design, the other for modelling and data-analysis. We split the activities into two groups, to better focus discussion for each of these templates.

-   [ ] <mark style="background: #FFF3A3A6;">Note: might need to reframe this and omit the field study stuff, but should probably state that we did this but decided not to pursue?</mark>

The objectives of the collaborative workshop activities were to aid in: 1. Understanding how modelling is done in practice to inform and tailor the preregistration templates to adequately capture key decision-points in the scientific modelling process; 2. Understanding how science in practice -- whether due to logistical, technical, feasibility or normative constraints -- differs from some idealised practice, in order to: 3. Identifying challenges and roadblocks to the use and uptake of preregistration in among ecological modellers, both in empirical and applied research contexts; and to 4. Using the above knowledge to inform the preregistration template and process design so they meet their intended purpose, and are adopted by our target user community for their particular use-cases.

### Workshop 1: Identifying Ecological Modelling Workflows, Norms, and Practice

The objective of this workshop was to identify modelling workflows that capture critical decision-steps in the model development cycle that should be included in a preregistration template. By "critical", we mean that without these decision-steps being included, the template would not adequately constrain researcher degrees of freedom nor b) transparently document the analytic decision-process. The activity consisted of two-parts, first, we individually reflected on our scientific process for a recent or memorable research project, listing the steps we undertook from the beginning to end. In the second part of this exercise, we came together as a group to collate each of the steps in our personal workflows onto pre-prepared workflow template (insert specific OSF link to the document, Supp Mat 1). The template consisted of several key 'phases' of the modelling process, with commonly implemented decision-steps grouped under each phase. The development of this modelling workflow sheet was derived from previous work by Gould et al. *in prep* (insert link to paper preprint) and is outlined in Appendix 1 ( #todo insert ). We sorted and categorised each decision-step in our personal workflows as belonging to a particular workflow phase. Next we grouped similar decision-steps that could be included as a single item on a preregistration template. At the end of this process we reviewed the suggested phases of the workflow to assess whether phases needed further discretisation, aggregation, removal, or amendment.

### Workshop 2: Understanding ecology- and modelling-specific challenges to adopting preregistration in research practice

This second activity aimed to identify critical challenges and roadblocks to undertaking preregistration in ecology and conservation in general, and for model-based research in particular. We facilitated a discussion structured around a series of pre-prepared questions and talking points informed by recent debates about preregistration in the literature ( #todo Supp Mat 1). These talking points were chosen with the aim of identifying problems in the content and or structuring of the phases and decision steps that were collectively identified in Workshop 1, as well as any conceptual or philosophical issues with preregistration in ecology, conservation, and in particular modelling, that might impact on the template structure, content and procedural implementation of preregistration. Discussion points included the following: - What should a research workflow look like for a modeller in an applied ecological setting who wishes to preregister their work? - What barriers are there for preregistering model-based research? - How do we accommodate the iterative cycle of model development into the preregistration process? - At what stage in a modelling or structured decision-making process should preregistration begin? - should we and how can we change the medium of the template and/or archiving platform to accomodate any roadblocks? - How could we change the medium of both the template and the archiving platform to accommodate these issues? - What procedural and or technical solutions could we implement to address these roadblocks to preregistration? - What decision-tools could be used to aid in completing a preregistration? - What sort of parsimony should we aim for in terms of the resolution of the templates and therefore their applicability across different methodologies?

### Analysis of Workshop Outputs & Design

> Analyse outputs of user-research workshops to inform the design of resource mockups

**1. Identify Scientific Workflows for ecological modelling, both real/in-practice and idealised.**

The goal of this analysis task was to describe a generalised workflow of the modelling cycle that will underpin the structure and content of the preregistration template. We used a combination of inductive and deductive coding ( #todo is this what it is called and #citation) using the qualitative coding software ATLAS.ti [-@atlasTI] to analyse the workshop outputs and generate a model of scientific workflow for model-based research that would constitute the presumed scientific workflow that would be used to guide the structure of the preregistration template. We collated and coded the workshop outputs using. All workshop outputs can be found here ( #todo Link to Workshop Outputs OSF repository), and the coded analysis can be found here ( #todo link to analysis files / screeen caps).

A preliminary idealised scientific workflow for ecological modelling research was synthesised after reviewing the literature on ‘good modelling practice’ and structured decision-making literature in ecological management. Steps in the workflow were coded and broken down into ‘phases’ consisting of ‘steps’ and ‘sub-steps’, in line with the ontology we specified earlier for translating a modelling workflow into a preregistration template (section link), <mark style="background: #FF5582A6;">(insert link to appendices or just present the final?)</mark>. Each decision-point in the personal modelling workflows generated from activity one of the collaborative workshop were coded as belonging to one of these phases, steps and sub-steps. We revised the coding structure (inductively is this the word?), according to patterns and themes identified across multiple personal workflows. This workflow is used to provide structure and content to the preregistration template ( #todo insert link to this workflow).

**2. Translating Workflows into Preregistration Templates**

We translated the model-based workflow identified in part one of workshop analysis into a text-based preregistration template. Reflecting on existing templates ( #todo link to OSF repository of templates page, and others, namely secondary data analysis, see Olmo's links), we identified common elements / units of preregistration templates that we mapped onto the generalised modelling workflow identified in part 1 of workshop outputs analysis. Following the features of a good guideline identified by Moallemi et al. [-@Moallemi2019], we provide examples of choices or illustrations to clarify each step of the preregistration, especially for phases or steps in the the modelling process that are semantically, epistemically and/or philosophically contentious or ambiguous (link to paper describing contentious issues in modelling history)\_. We synthesised the following ontology with which to build our preregistration template: 1.        Decision Phase - each has a “title” and “description” or “definition”, steps are grouped under phases. (Note, that these are being scrapped from the template and are turned into tool-tips for formatting) 2.        Decision Step - each has a “title” and “description” or “definition” choices are grouped into ‘steps’, which corresponds to an activity or set of related activities in a scientific workflow. (again, these are being scrapped and turned into tool-tips. CHECK THIS, it's changed) 3.        Choices —\> each ‘choice’ corresponds to a uniquely numbered item on the PRT requiring a response from user. The style should be in the form of a directive or a question. E.g. “explain how you will do xyzzy”. Or “What performance measure will you use for assessing ‘goodness of fit’?”

**3. Identifying challenges and solutions for implementing preregistration in model-based research**

We analysed the outputs of Activity 2 in the collaborative workshop identifying barriers and solutions to undertaking preregistration in model-based research and to broader adoption of preregistration in the modelling community of practice. The analysis of Activity 2 in the collaborative workshop was used to inform the design of a suitable preregistration procedure for preregistering MBR in the domain of ecology.

## User Testing

> Build mock-ups of the product and test with users

While user-testing is an essential step in the UCD framework, we also concur with @Devezer2020 that open-science reforms should be evaluated against their stated objectives. We therefore preregistered a real-world model-based research problem as a case-study for evaluating and testing the preregistration templates and proposed adaptive preregistration methodology.

### Case Study

We preregistered a real-world study of vegetation responses to environmental flows in Victoria, Australia. This study is one component of a larger set of studies within the Stage 6 Victorian Environmental Flows Monitoring and Assessment Program (VEFMAP), managed by Department of Energy, Environment and Climate Action (DEECA) and delivered through the Arthur Rylah Institute for Environmental Research. The case-study analysis aimed to evaluate the effectiveness of environmental-flows on riparian vegetation with the following objectives: - Quantify short and medium-term vegetation responses to individual flow events and regimes.  - Describe detailed vegetation trends associated with season, flow, rainfall and other factors such as livestock grazing and exotic vegetation  - Quantify the effect of various factors on vegetation in order to guide future management expectations and planning.

Details of the study background, methods and results are described in Appendix A4.

### Evaluation Criteria & Rationale

-   [ ] *Evaluation Criteria and Rationale (should this be moved to earlier aims section?* Seems like there's a little bit of overlap between the two...

The content of the template should be 'parsimonious' in that the included items within the template should adequately a) constrain researcher degrees of freedom and by being specific, precise and exhaustive [@Wicherts2016] b) facilitate transparent documentation of the modelling process and analytic decisions during model development and analysis, while c) not being too detailed to the point that the preregistration is too difficult and time-consuming to complete.

Many modellers [e.g. @MacEachern2019] #todo #citation are concerned about preregistration making modelling a more difficult task than it already is, and hampering the ability to meet project deadlines, especially in applied settings in which models directly underpin decisions. Modellers already struggle with sufficiently documenting models during publication and reporting , and there is a resource cost to increasing transparency of model-reporting [@deVos:2011uk]. Consequently we will also assess the preregistration template and process on its perceived feasibility and practicality.

### Evaluation Method

Rather than attempting to reconstruct the process of creating the preregistration through using interview questions alone and digging back through the version history of a completed preregistration [e.g. @Pu2019], we aimed to capture the analysts 'thinking in the moment' by utilising the version-control and collaborative project management features of Git and GitHub (Ben, Smith, and Bailey 2018; Blischak, Davenport, and Wilson 2016) as a tool for documenting both the analytic decisions of the researcher, and for using the preregistration process itself to ‘live-develop’ the template template. For example, if it turned out that we had missed an important item, or perhaps the phrasing of an item or the order and structure of the template needed to change whilst in the process of completing the preregistration, the suggested change, and its justification would be both recorded in GitHub and linked to the actual preregistration.

Finally, we supplemented this ‘in situ’ testing and evaluation process with follow-up semi-structured interviews on 9th May 2024 in order to capture more detailed reflections of the analysts. Lead analysts of the case study were often under strict time constraints for meeting deliverables on the project, and the process of preregistration was a new and unfamiliar task, so capturing reflections on the broader preregistration process during preregistration and analysis of the case-study was not always possible.

## Implementation

> Implement the product, write technical implementations and build the product

We wrote the initial template as a text-based document in Rmarkdown. The initial template used to preregister the case-study underwent a series of changes throughout the preregistration process, including changes to the organisation, content, help-text and visual appearance. The final preregistration template is made available as a quarto markdown extension ( #citation & link to doc) that can be installed and used to populate a new preregistration document for a new modelling study, and as a form that can be used with the `preregr` R package [@Gjalt-Jorn2023].

To aid the case-study analysts in using the adaptive preregistration methodology to preregister their study, we wrote a detailed user-guide outlining the key principles of adaptive preregistration and a protocol for implementing adaptive preregistration using git and GitHub. The guide is presented as a publicly available website and also includes the template.

@@

## User-research and analysis

The outputs and analysis for both Workshop 1 and 2 is archived at (insert OSF link). The modelling workflow generated from the analysis of Workshop 1 that informed the development of the initial preregistration template is archived at the same link. During user-testing with our case-study, the preregistration template structure, and therefore its underlying modelling workflow on which it is based underwent, a series of changes. Consequently we present the revised workflow alongside the refined preregistration template in section X below.

Discussions from Workshop Activity 2 focussing on challenges to preregistering MBR revealed that the iterative nature of model development is a significant barrier to adopting preregistration for model-based research. Modellers felt that adopting preregistration in their workflows was impossible because preregistration was understood to be a process that precludes iterative model development because its inner logic presumes a linear research workflow premised on hypothesis-testing research, and because it requires researchers to confirm an unalterable version of the analysis plan prior to collecting or analysing the data [@Pu2019; @mertens2019].

The key tenets of preregistration are based on NHST-focused accounts of questionable research practices that prescribe the principle of 'decision-independence' where analysis decisions are made independently of and *a priori* to analysing the data [@Srivastava2018; @Gelman2013]. However, analysts reflected on their personal modelling workflows and described situations in the modelling process that may violate the principle of 'decision-independence' that could not be avoided and were considered to be an intrinsic feature of the model-development process. During modelling a researcher will almost always engage in state-dependent or data-dependent decision-making -- that is, decisions about whether to alter the model in a subsequent round of model development depend on the results of model testing, evaluation and analysis at the end of the previous iteration of model development.

Very often in modelling, there are decisions that are necessarily dependent on the outcome of previous analytic decisions in the modelling workflow [@Liu2020]. Firstly, preliminary or investigatory analyses might need to be conducted before being able to specify future decision-steps in the analysis plan. For example, modellers might need to check the distribution of particular variables in order to determine how to specify the model, perform assumption checks, or check for collinearity or spatial autocorrelation. Secondly, data-dependent decisions from model checking may justifiably result in changes to either the analysis or to interpretations of the model [@Srivastava2018]. For instance, inspection of the residuals and other model checking tests may force the modeller to return to earlier decision-points and change the planned analysis or model. Other times, the planned analysis and specified model may simply not converge, or the model is saturated and runs out of variation to apportion such that the model and or the model fitting algorithm must be respecified and re-implemented. Other times, the modelling process itself may generate knowledge and learning about the system, and the model structure changes throughout the process of model development.

Data-dependent decision-making is not necessarily intrinsically *questionable*, but it does have the *potential* to reduce the reliability of the model and its findings if they are exploited in service of artificially inflating the accuracy or precision of the model, its predictions or evaluation tests to the effect that the model is perceived to be more credible than it would be if that practice did not occur -- essentially data-dependent decisions that lead to consumers of the model placing unsubstantiated belief in the reliability, validity and utility of the model and its outputs are questionable [@Gould:2019ebg] and can be mitigated through the practice of preregistration.

The concept of preregistering all modelling choices *a priori* was seemingly overwhelming for analysts. They described situations where decisions are too difficult to anticipate, or cannot be made in advance, particularly for decisions occurring at the later phases of model construction and development process (Figure 2, TBC), where downstream decisions depend on the outcomes of earlier decisions and outcomes of modelling analyses. Some decisions might not be able to be specified on the first attempt at writing the preregistration until the model is fully or at least close to fully specified, for example, specifying precisely what and how sensitivity analyses or uncertainty analyses will be conducted.

-   Finally, sometimes there are some decisions that are too difficult to anticipate, or simply cannot be made in advance [@Srivastava2018].
-   I think Pu or Pavelin describe the same thing... they coin a name for these types of decisions too but I can't find it.

## Design

### Adaptive Preregistration: A methodology for preregistering model-based research
